---
title: "Covid19_Report"
author: "C. Mayr"
date: "2024-05-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal of this Report
In this report I will download data about covid cases in the US. 
My goal is to figure out if states with higher population density had more
cases per person in this state or not.


## Loading the data

First step is to include the libraries tidyverse and lubridate. We will need 
them later in the project.

```{r IncludeLibraries, message=FALSE}
library(tidyverse)
library(lubridate)
```


Next, load the data from the website, as shown in the lecture. For the analysis
we will only need the data sets cases and deaths.


```{r LoadData, echo=TRUE, message=FALSE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/" %>%
          str_c("master/csse_covid_19_data/csse_covid_19_time_series/")

filenames <- c("time_series_covid19_confirmed_US.csv",
"time_series_covid19_confirmed_global.csv",
"time_series_covid19_deaths_US.csv",
"time_series_covid19_deaths_global.csv",
"time_series_covid19_recovered_global.csv")

urls <- str_c(url_in, filenames)

#Load relevant US data in Tables
US_cases <- read_csv(urls[1])
US_deaths <- read_csv(urls[3])

```


## Prep the data


Here the columns with the dates and cases will be pivoted into rows. We 
eliminate the columns Lat and Long and convert the date in to the 
date datatype.
At last, the two tables will be fully joined into one.


```{r PrepData, echo=FALSE, message=FALSE}

US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
  
US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))


US <- US_cases %>% 
        full_join(US_deaths)


```

In the next step the cases and deaths will be summarized by state. This has to
happen in two steps. First we take the maximum value of cases, deaths and 
population and group by county. Afterwards we summarize the numbers for each
county and group by state.
I save the total population of the USA in the variable total_population. 
This was a verfication step for myself, after some trouble with the numbers.

```{r PrepData2, echo=FALSE, message=FALSE}

#First summarize the cases and deaths per county and take the max population. 
# This step is necessary to get the right population in the next step
US_by_State_totals <- US %>%
    group_by(Admin2, Province_State, Country_Region) %>%
    summarize(cases = max(cases), deaths = max(deaths), 
              Population = max(Population)) %>% 
  select(Admin2, Province_State, Country_Region, cases, deaths, Population) %>%
    ungroup() %>% filter(Population > 0)

# Group the data to get the numbers per state
US_by_State_totals <- US_by_State_totals %>%
    group_by(Province_State, Country_Region) %>%
    summarize(cases = sum(cases), deaths = sum(deaths), 
              Population = sum(Population)) %>% 
  select(Province_State, Country_Region, cases, deaths, Population) %>%
    ungroup() %>% filter(Population > 0)
    
# Check if summarizing was correct
total_population <- US_by_State_totals %>%
  summarize(total_population = sum(Population, na.rm = TRUE))


```

To calculate the population density in each state, found a source on github
with the areas of each state. 
I will download this data set, rename the columns, and join the area column 
to the existing table.
Furthermore I mutate the table with the calculated population desity and
the cases per person per hundred per state. (This just gives a nicer number
to compare)

```{r PrepAreaData, echo=TRUE, message=FALSE}

uid_lookup_url <- "https://raw.githubusercontent.com/" %>%
                    str_c("jakevdp/data-USstates/master/state-areas.csv")

AreaData <- read_csv(uid_lookup_url) 

AreaData <- AreaData %>%
  rename(Province_State = state,
         area = "area (sq. mi)")
      
US_Cov_Area <- US_by_State_totals %>% 
        full_join(AreaData) %>% filter(area > 0)

US_Cov_Area <- US_Cov_Area %>%
    mutate(PopulationDensity = Population/area,
           CasesPerHundred = (cases*100)/Population)


```


## Visualizing Data

To get an first impression on how the data looks and to compare the density data
with the area data, I decided to create bar charts for each. 
I don't want to overload the charts. Therefore, I decided just to display the 
top ten in each category. 
If my hypothesis is right, the states with the highest density should also be
in the top ten of the most cases.
Unfortunately this looks not like if it would be the case.


```{r VizData, echo=TRUE, message=FALSE}


top_10_states_Dens <- US_Cov_Area %>%
  arrange(desc(PopulationDensity)) %>%
  slice_head(n = 10)

states <- top_10_states_Dens$Province_State
densitys <- top_10_states_Dens$PopulationDensity


par(mfrow = c(1, 2))

barplot(densitys, 
        names.arg = states, 
        las = 2,  # Rotate the state names for better readability
        col = "steelblue", 
        main = "Top 10 highest density", 
        xlab = "", 
        ylab = "Persons per sqaremile",
        cex.names = 0.7)
  

top_10_states_cases <- US_Cov_Area %>%
  arrange(desc(CasesPerHundred)) %>%
  slice_head(n = 10)

states <- top_10_states_cases$Province_State
cases <- top_10_states_cases$CasesPerHundred

barplot(cases, 
        names.arg = states, 
        las = 2,  # Rotate the state names for better readability
        col = "red", 
        main = "Top 10 States most cases", 
        xlab = "", 
        ylab = "cases per hundred people",
        cex.names = 0.7)

par(mfrow = c(1, 1))

```



## Data Model

To get a better overview over all of the states, I implemented a linear model, 
which predicts the cases over the area. 


```{r DensPop_model, echo=TRUE,}

mod <- lm(CasesPerHundred ~ PopulationDensity, data = US_Cov_Area)

CasesPerSquaremile_withPred <- US_Cov_Area %>% 
  mutate(pred = predict(mod))

CasesPerSquaremile_withPred %>% 
  ggplot() + 
  geom_point(aes(x= PopulationDensity, y= CasesPerHundred, color = "cases")) + 
  geom_point(aes(x= PopulationDensity, y= pred, color = "prediction")) + 
  theme(legend.position="bottom", axis.text.x = element_text(angle = 90)) + 
  labs(title = "Cases over area with prediction model", y = NULL)


```


## Conclusion

By just looking at the data points from the predicted model, it seems like an 
indirect proportional relationship between the two variables.
By looking at the data points one can clearly say there is no correlation 
between them.

Finally I want to talk about the biases in the data I used to make the analysis.

The data is collected form confirmed covid cases. A lot of people were involved
to collect this data and people make mistakes.Furthermore, what is about the
unconfirmed cases. For the analysis they would be very interesting and also 
important to know. I maybe also made a mistake by using the max value and not
the increasing rate per day.

The population data can also be biased. In the population data are "only" 
registered people that live in that area. What about unregistered people. This 
could also shift the predicted data. 










